From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-19 Pi+/paper/finite.tex

Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots

2. In the usual setting, there are are many relations that are true for the elements of the group. For example, in the
(Z mod 3), we have a relation 1 + 1 + 1 = 0.

3. A group that has no other relations than the ones directly implied by
the definition of a group is called free.

4. A group is said to be generated by a set of its elemenst (e1, ..., en)
if every element in the group can be constructed from these elements
composing operations of + and inverse. For example, Z is a free group,
generated by one-element set (1).
4a. Thus, a free group g can be thought as sequences of letters from the
generating set, divided by the congruence closure of the group relation.
Definition: Free group.

5. One useful way of defining groups is by their presentations.
Presentation of a group builds a group by specifying what is the set of
generators - thus making a free group - a collection of equations that have
to be satisfied in the resulting group. For example, we could take the
generators to be the set (1) (giving out Z), and then add an equation (1 +
1 + 1 = 0) - thus, getting (Z mod 3).

5a. The generatos can be thought of as primitive operations in a language,
and the group structure gives us the way of composing these operations and
inverting them.

6. Building a group like that is possible because of the universal property
of the free group.
Theorem: Universal property of a free group.
Definition: formal definition of a free group.

7. Why do it like that? Because it is a very syntactic approach to doing
the theory. The way described previously operates directly in the semantics
of the group operation - equality of elements (such as the result of adding
two elements) has to be computed externally.
Here, the equality can be computed using a rewriting system specified by
the group axioms and the additional equations provided. However, because
equations are not directed, it is not always possible to construct a
well-behaving rewriting system. In general, the word problem (i.e. deciding
if two words are equal) is proven to be undecidable.

\subsection{Permutation Groups}

8. The group that we are interested in is the group of permutations on a
fininte set. We can define it formally in two different ways:
8a. Taking the type Aut (Fin n) and showing that it satisfies the group
axiom.
8b. Writing up a presentation - choosing generating set and equations that
have to be satisfied.

9. By proving these two to be equivalent, we are bridging the gap between
syntactic and semantic side in the whole proof - by assinging the meaning
(a bijection) to a word (a program).
9a. However, proving them equivalent will require, in particular, solving
the word problem (in this specific instance).

@@

\begin{definition}
  Let $A$ be a type, and $\List[A]$ the free monoid on $A$. The free group $F(A)$ on $A$ is the set-quotient of
  $\List[A + A]$ by the congruence closure of the relation $(\inl(a) \cons \inr(a) \cons \nil) \sim \nil$.
\end{definition}

\begin{proposition}
  Let $\eta_{A} : A \to F(A)$ be given by $a \mapsto \inl(a) \cons \nil$. Given any group $G$ with a map $f : A \to G$,
  there is a canonical lifting of $f$ to a group homomorphism $\extend{f} : F(A) \to G$, such that
  $\extend{f} \comp \eta_{A} \htpy f$, that is, the type of group homomorphisms $h : \Grp(F(A),G)$ satisfying
  $h \comp \eta_{A} \htpy f$ is contractible. Equivalently, composition by $\eta_{A}$ is an equivalence
  $\Grp(F(A),G) \eqv (A \to G)$.
\end{proposition}

\todo{diagram}
\todo{Plus/Minus}

\begin{definition}
  Let $A$ be a type and $R : A \to A \to \UU$ a binary relation on $A$. The group $G$ presented by
  $\langle A ; R \rangle$ is given by the set-quotient of the free group $F(A)$ by the normal closure $N_{R}$ of $R$, or
  equivalently, as the coequaliser
  \[\begin{tikzcd}
      FR && FA && G
      \arrow[shift right=2, from=1-1, to=1-3]
      \arrow[shift left=2, from=1-1, to=1-3]
      \arrow[two heads, from=1-3, to=1-5]
    \end{tikzcd}\]
\end{definition}

\todo{Universal property}
\todo{Examples: empty relation, van Kampen of $\pi_{1}$}

>>>>

\subsection{Free groups}

Given a generating set $A$, we can naively construct a \emph{free group on $A$},
whose elements are drawn from the alphabet $A$, and closed under the group
operations of multiplication and inverse, identified by the group axioms.
For example, the singleton set generates the additive group of integers,
$\mathbb{Z}$.

Usually, there are are many equations, besides the group axioms, that hold for
the elements of the group. For example, in the group $\mathbb{Z}/3\mathbb{Z}$,
we have an equation $1 + 1 + 1 = 0$. A free group has the property that has no
other relations than the ones directly implied by the group axioms.

\todo{What's the best notation for HITs?}
\begin{definition}
  Given an $\hSet$ $A$, the free group $F(A)$ on it is given by a higher
  inductive type with the following point and path constructors. Notice the
  similarity with the definition of a group structure in~\ref{subsec:groups},
  but note that each operation here is a generator for the type $F(A)$\dots.
  \begin{itemize}
    \item An inclusion function $\eta_{A} : A \to F(A)$
    \item A multiplication function $m : F(A) \times F(A) \to F(A)$
    \item An element $e : F(A)$
    \item An inverse function $i : F(A) \to F(A)$
  \end{itemize}
  \smallskip
  \begin{itemize}
    \item For every $x, y, z : F(A)$, a path $\term{assoc} : m(x, m(y, z)) \id m(m(x, y), z)$
    \item For every $x : F(A)$, paths $\term{unitr} : m(x, e) \id x$ and $\term{unitl} : m(e, x) \id x$
    \item For every $x : F(A)$, paths $\term{invr} : m(x, i(x)) \id e$ and $\term{invl} : m(i(x), x) \id e$
    \item A 0-truncation, for every $x, y : F(A)$ and $p, q : x \id y$, a 2-path $\term{trunc} : p \id q$
  \end{itemize}
\end{definition}

A group homomorphism between groups is a function between the underlying sets
that preserves the group structure, which we write as $G_1 \to^G G_2$. With
this, we can state the universal property of free groups.

\begin{proposition}[Universal Property of $F(A)$]~\label{prop:free-groups}
  Given a group $G$ and a map $f : A \to G$, there is a unique group
  homomorphism $\extend{f} : F(A) \to^G G$ such that $\extend{f} \comp \eta_A
  \htpy f$. Equivalently, composition with $\eta_A$ gives an equivalence $F(A)
  \to^G G \eqv A \to G$. Alternatively, the type of group homomorphisms $h :
  F(A) \to^G G$ satisfying $h \comp \eta_A \htpy f$ is contractible.

  % https://q.uiver.app/?q=WzAsMyxbMCwyLCJBIl0sWzAsMCwiRihBKSJdLFsyLDAsIkciXSxbMCwxLCJcXGV0YV9BIl0sWzAsMiwiZiIsMl0sWzEsMiwiXFxleHRlbmR7Zn0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkYXNoZWQifX19XSxbMyw0LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}
	{F(A)} && G \\
	\\
	A
	\arrow[""{name=0, anchor=center, inner sep=0}, "{\eta_A}", from=3-1, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "f"', from=3-1, to=1-3]
	\arrow["{\extend{f}}", dashed, from=1-1, to=1-3]
	\arrow["\id", Rightarrow, draw=none, from=0, to=1]
\end{tikzcd}\]
\end{proposition}

This definition of $F(A)$ however has lots of path constructors corresponding to
each group axiom. Alternatively, the free group construction can be thought of
as drawing letters from the generating set, while adding formal inverses, and
constructing words which are sequences of these letters.

Each element in the sequence can either be an element drawn from the generating
set $a$, or an inverse of such an element $\inv{a}$. If we take the disjoint
union of $A$ with itself, that is, $A + A$ as the underlying set, we can use
$\inl/\inr$ to mark the elements. Hence, we can encode the free group using the
free monoid, that is, lists.

\begin{definition}
  Let $A$ be an $\hSet$, and $\List[\blank]$ the free monoid. The free group
  $F(A)$ on $A$ is the set-quotient of $\List[A + A]$ by the congruence closure
  of the relation $a \cons \inv{a} \cons \nil \sim \nil$.
\end{definition}

\begin{proposition}
  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the
  empty list $\nil$ for the neutral element, multiplication given by list append
  $\append$, and inverse given by list reversal. Further, $F(A)$ with $\eta_A :
  A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal property of free
  groups.~\cref{prop:free-groups}
\end{proposition}

\subsubsection{Group presentations}

One useful way of defining groups is by their presentations. A presentation of a
group builds it by starting from the free group $F(A)$ and introducing a
collection of equations that have to be satisfied in the resulting group. For
example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 + 1
= 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$.

The generators can be thought of as primitive operations in a (reversible)
programming language, group structure gives the way of composing these
operations and inverting them, and relations describe how these primitive
operations interact with each other.

Instead of opereating directly in the semantics of the group operation, here the
focus is on the syntax. While before, the equality of elements (such as the
result of multiplication of two elements) had to be computed externally, now it is
reduced to a \emph{word problem}, i.e., deciding one word can be reduced to another
using group's equations.

However, because equations are not directed, is not always possible to construct a
well-behaving rewriting system. In general, the word problem is proven to be undecidable.

\jk{Example?}

\begin{definition}
  Let $A$ be an $\hSet$ and $R : (A + A) \to (A + A) \to \UU$ a binary relation
  on $A + A$. The group $G$ presented by $\langle A ; R \rangle$ is given by the
  set-quotient of the free group $F(A)$ by the closure of $R$, or equivalently,
  as the coequaliser
  \[\begin{tikzcd}
      FR && FA && G
      \arrow[shift right=2, from=1-1, to=1-3]
      \arrow[shift left=2, from=1-1, to=1-3]
      \arrow[two heads, from=1-3, to=1-5]
    \end{tikzcd}\]
\end{definition}

\vc{do we need this level of detail?}
\todo{Universal property}
\todo{Examples: empty relation, van Kampen of $\pi_{1}$}

The above definition is correct because the universal property of the free group allows 
for properly extending the relation on the generating set to the whole group.
\jk{Turn that into a propostion?}

\subsection{Permutation groups}

Because $\Pi$ The group that we are interested in is the group of permutations on a
fininte set. We can define it formally in two different ways:
8a. Taking the type Aut (Fin n) and showing that it satisfies the group
axiom.
8b. Writing up a presentation - choosing generating set and equations that
have to be satisfied.

9. By proving these two to be equivalent, we are bridging the gap between
syntactic and semantic side in the whole proof - by assinging the meaning
(a bijection) to a word (a program).

@@
